Estimating carrying capacity for juvenile salmon using quantile random forest models

### 2020-12-04 analysis/paper/paper.tex
Implemented revisions suggested by Mike, Sarah and Richie before final submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models},
  pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
  pdfkeywords={carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

@@

\title{Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models}
\author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
\date{December 02, 2020}

\begin{document}
\maketitle
\begin{abstract}
Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
\end{abstract}

@@

The decline of anadromous Pacific salmonids (\emph{Oncorhynchus spp.}) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H's: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids (Bond et al. 2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (\emph{O. mykiss}) in the Pacific Northwest, USA (NOAA Fisheries 2016a, 2016b). Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge (Bond et al. 2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@@

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor proxy of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) (Meinshausen 2006) models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling (Koenker and Bassett Jr 1978, Cade and Noon 2003) with the flexibility and framework of random forest models (Breiman 2001). They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts (Knudby et al. 2010). Quantile random forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers (Prasad et al. 2006). In addition, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors (Terrell et al. 1996, Cade and Noon 2003).

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened (NOAA Fisheries 2016a, 2016b). Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

\hypertarget{study-site}{%
\subsection{Study Site}\label{study-site}}

Habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \ref{fig:wtsd-map}). The Columbia River basin covers more than 668,000 km\(^2\) draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) (Volk et al. 2017) and were downloaded from \url{https://www.champmonitoring.org}. Data from the following eleven CHaMP watersheds were used in this study: Asotin, Entiat, John Day, Lemhi, Methow, Minam, South Fork Salmon, Tucannon, Upper Grande Ronde, Wenatchee and Yankee Fork. Juvenile density and abundance data were collected in a subset of seven watersheds (see Table \ref{tab:fish-hab-sites} and Figure \ref{fig:wtsd-map}), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project (Volk et al. 2017).

@@

\hypertarget{habitat-covariate-selection}{%
\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection}}

@@

\hypertarget{qrf-model-fit}{%
\subsection{QRF Model Fit}\label{qrf-model-fit}}

@@

\hypertarget{extrapolating-to-other-sites}{%
\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites}}

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from (Larsen et al. 2008, 2016). Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification (Whittier et al. 2011), the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \ref{tab:gaa-covars}). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the \emph{svyglm} function from the \emph{survey} package (Lumley 2020) in R software (R Core Team 2019). The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients (Nahorniak et al. 2015). We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.

@@

\hypertarget{model-validation}{%
\subsection{Model Validation}\label{model-validation}}

@@

\hypertarget{habitat-covariate-selection-1}{%
\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection-1}}

@@

\hypertarget{qrf-model}{%
\subsection{QRF Model}\label{qrf-model}}

@@

\hypertarget{extrapolating-to-other-sites-1}{%
\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites-1}}

@@

\hypertarget{model-validation-1}{%
\subsection{Model Validation}\label{model-validation-1}}

@@

\hypertarget{a-tool-to-estimate-habitat-capacity}{%
\subsection{A Tool to Estimate Habitat Capacity}\label{a-tool-to-estimate-habitat-capacity}}

@@

\hypertarget{biological-expectations-from-qrf-model}{%
\subsection{Biological Expectations from QRF Model}\label{biological-expectations-from-qrf-model}}

@@

\hypertarget{extrapolation-model}{%
\subsection{Extrapolation Model}\label{extrapolation-model}}

@@

\hypertarget{the-future-improving-habitat-data}{%
\subsection{The Future: Improving Habitat Data}\label{the-future-improving-habitat-data}}

@@

\hypertarget{conclusions-and-next-steps}{%
\subsection{Conclusions and Next Steps}\label{conclusions-and-next-steps}}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allen2000}{}%
Allen, M. A. 2000. Seasonal microhabitat use by juvenile spring chinook salmon in the yakima river basin, washington. Rivers 7:314--332.

@@

\leavevmode\hypertarget{ref-Bond2018}{}%
Bond, M. H., T. G. Nodine, T. J. Beechie, and R. W. Zabel. 2018. Estimating the benefits of widespread floodplain reconnection for columbia river chinook salmon. Canadian Journal of Fisheries and Aquatic Sciences:1--15.

\leavevmode\hypertarget{ref-Braun2011}{}%
Braun, D. C., and J. D. Reynolds. 2011. Relationships between habitat characteristics and breeding population densities in sockeye salmon (oncorhynchus nerka). Canadian Journal of Fisheries and Aquatic Sciences 68:758--767.

@@

\leavevmode\hypertarget{ref-Brett1952}{}%
Brett, J. R. 1952. Temperature tolerance in young pacific salmon, genus oncorhynchus. Journal of the Fisheries Board of Canada 9:265--323.

@@

\leavevmode\hypertarget{ref-Carle1978}{}%
Carle, F. L., and M. R. Strub. 1978. A new method for estimating population size from removal data. Biometrics:621--630.

\leavevmode\hypertarget{ref-Carmichael2020}{}%
Carmichael, R. A., D. Tonina, E. R. Keeley, R. M. Benjankar, and K. E. See. 2020. Some like it slow: A bioenergetic evaluation of habitat quality for juvenile chinook salmon in the lemhi river, idaho. Canadian Journal of Fisheries and Aquatic Sciences.

@@

\leavevmode\hypertarget{ref-CHaMP2016}{}%
CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.

\leavevmode\hypertarget{ref-Chapman1951}{}%
Chapman, D. G. 1951. Some properties of the hypergeometric distribution with applications to zoological sample censuses. University of California Press.

\leavevmode\hypertarget{ref-Compton2012}{}%
Compton, T. J., M. A. Morrison, J. R. Leathwick, and G. D. Carbines. 2012. Ontogenetic habitat associations of a demersal fish species, pagrus auratus, identified using boosted regression trees. Marine Ecology Progress Series 462:219--230.

\leavevmode\hypertarget{ref-Cramer2009}{}%
Cramer, S. P., and N. K. Ackerman. 2009. Linking stream carrying capacity for salmonids to habitat features. Pages 225--254 \emph{in} American fisheries society symposium.

\leavevmode\hypertarget{ref-Dodrill2016}{}%
Dodrill, M. J., and C. B. Yackulic. 2016. Nonlinear relationships can lead to bias in biomass calculations and drift-foraging models when using summaries of invertebrate drift data. Environmental biology of fishes 99:659--670.

@@

\leavevmode\hypertarget{ref-Eastwood2003}{}%
Eastwood, P. D., G. J. Meaden, A. Carpentier, and S. I. Rogers. 2003. Estimating limits to the spatial extent and suitability of sole (solea solea) nursery grounds in the dover strait. Journal of Sea Research 50:151--165.

\leavevmode\hypertarget{ref-Fausch1988}{}%
Fausch, K. D., C. L. Hawkes, and M. G. Parsons. 1988. Models that predict standing crop of stream fish from habitat variables: 1950-85. Notes.

@@

\leavevmode\hypertarget{ref-Hillman1987}{}%
Hillman, T. W., J. Griffith, and W. Platts. 1987. Summer and winter habitat selection by juvenile chinook salmon in a highly sedimented idaho stream. Transactions of the American Fisheries Society 116:185--195.

@@

\leavevmode\hypertarget{ref-Holecek2009}{}%
Holecek, D. E., K. J. Cromwell, and B. P. Kennedy. 2009. Juvenile chinook salmon summer microhabitat availability, use, and selection in a central idaho wilderness stream. Transactions of the American Fisheries Society 138:633--644.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Estimating carrying capacity for juvenile salmon using quantile random forest models},
  pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
  pdfkeywords={carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

@@

\title{Estimating carrying capacity for juvenile salmon using quantile random forest models}
\author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
\date{December 03, 2020}

\begin{document}
\maketitle
\begin{abstract}
Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (\emph{Oncorhynchus} spp.). A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest model to provide estimates of habitat carrying capacity for Chinook salmon (\emph{O. tshawytscha}) parr during the summer months, at both the site and watershed scale. Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
\textbf{Key words:} carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
\end{abstract}

@@

The decline of anadromous Pacific salmon (\emph{Oncorhynchus} spp.) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H's: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids (Bond et al. 2019, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (\emph{O. mykiss}) in the Pacific Northwest, USA (NOAA Fisheries 2016a, 2016b). Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge (Bond et al. 2019, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@@

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor proxy of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) (Meinshausen 2006) models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling (Koenker and Bassett Jr. 1978, Cade and Noon 2003) with the flexibility and framework of random forest models (Breiman 2001). They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts (Knudby et al. 2010). Quantile random forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers (Prasad et al. 2006). In addition, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors (Terrell et al. 1996, Cade and Noon 2003).

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter), spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened (NOAA Fisheries 2016a, 2016b). Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

\hypertarget{study-site}{%
\subsection{Study site}\label{study-site}}

Habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \ref{fig:wtsd-map}). The Columbia River basin covers more than 668,000 km\(^2\) draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) (Volk et al. 2017) and were downloaded from \url{https://www.champmonitoring.org}. Data from the following eleven CHaMP watersheds were used in this study: Asotin, Entiat, John Day, Lemhi, Methow, Minam, South Fork Salmon, Tucannon, Upper Grande Ronde, Wenatchee and Yankee Fork Salmon. Juvenile density and abundance data were collected in a subset of seven watersheds (see Table \ref{tab:fish-hab-sites} and Figure \ref{fig:wtsd-map}), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project (Volk et al. 2017).

@@

\hypertarget{habitat-covariate-selection}{%
\subsection{Habitat covariate selection}\label{habitat-covariate-selection}}

@@

\hypertarget{qrf-model-fit}{%
\subsection{QRF model fit}\label{qrf-model-fit}}

@@

\hypertarget{extrapolating-to-other-sites}{%
\subsection{Extrapolating to other Sites}\label{extrapolating-to-other-sites}}

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from (Larsen et al. 2008, 2016). Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification (Whittier et al. 2011), the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \ref{tab:gaa-covars}). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the \emph{svyglm} function from the \emph{survey} package (Lumley 2020) in R software (R Core Team 2019). The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients (Nahorniak et al. 2015). We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so some form of spatial averaging of capacity predictions to generate larger scale capacity estimates is necessary.

@@

\hypertarget{model-validation}{%
\subsection{Model validation}\label{model-validation}}

@@

\hypertarget{habitat-covariate-selection-1}{%
\subsection{Habitat covariate selection}\label{habitat-covariate-selection-1}}

@@

\hypertarget{qrf-model}{%
\subsection{QRF model}\label{qrf-model}}

@@

\hypertarget{extrapolating-to-other-sites-1}{%
\subsection{Extrapolating to other sites}\label{extrapolating-to-other-sites-1}}

@@

\hypertarget{model-validation-1}{%
\subsection{Model validation}\label{model-validation-1}}

@@

\hypertarget{a-tool-to-estimate-habitat-capacity}{%
\subsection{A tool to estimate habitat capacity}\label{a-tool-to-estimate-habitat-capacity}}

@@

\hypertarget{biological-expectations-from-qrf-model}{%
\subsection{Biological expectations from QRF model}\label{biological-expectations-from-qrf-model}}

@@

\hypertarget{extrapolation-model}{%
\subsection{Extrapolation model}\label{extrapolation-model}}

@@

\hypertarget{the-future-improving-habitat-data}{%
\subsection{The future: improving habitat data}\label{the-future-improving-habitat-data}}

@@

\hypertarget{conclusions-and-next-steps}{%
\subsection{Conclusions and next steps}\label{conclusions-and-next-steps}}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allen2000}{}%
Allen, M. A. 2000. Seasonal microhabitat use by juvenile spring Chinook salmon in the Yakima River basin, Washington. Rivers 7:314--332.

@@

\leavevmode\hypertarget{ref-Bond2019}{}%
Bond, M. H., T. G. Nodine, T. J. Beechie, and R. W. Zabel. 2019. Estimating the benefits of widespread floodplain reconnection for Columbia River Chinook salmon. Canadian Journal of Fisheries and Aquatic Sciences 76:1212--1226.

\leavevmode\hypertarget{ref-Braun2011}{}%
Braun, D. C., and J. D. Reynolds. 2011. Relationships between habitat characteristics and breeding population densities in sockeye salmon (\emph{Oncorhynchus nerka}). Canadian Journal of Fisheries and Aquatic Sciences 68:758--767.

@@

\leavevmode\hypertarget{ref-Brett1952}{}%
Brett, J. R. 1952. Temperature tolerance in young pacific salmon, genus \emph{oncorhynchus}. Journal of the Fisheries Board of Canada 9:265--323.

@@

\leavevmode\hypertarget{ref-Carle1978}{}%
Carle, F. L., and M. R. Strub. 1978. A new method for estimating population size from removal data. Biometrics 34:621--630.

\leavevmode\hypertarget{ref-Carmichael2020}{}%
Carmichael, R. A., D. Tonina, E. R. Keeley, R. M. Benjankar, and K. E. See. 2020. Some like it slow: A bioenergetic evaluation of habitat quality for juvenile Chinook salmon in the Lemhi River, Idaho. Canadian Journal of Fisheries and Aquatic Sciences 77:1221--1232.

@@

\leavevmode\hypertarget{ref-CHaMP2016}{}%
CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the Columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.

\leavevmode\hypertarget{ref-Chapman1951}{}%
Chapman, D. G. 1951. Some properties of the hypergeometric distribution with applications to zoological sample censuses. University of California Press, Berkeley, California, USA.

\leavevmode\hypertarget{ref-Compton2012}{}%
Compton, T. J., M. A. Morrison, J. R. Leathwick, and G. D. Carbines. 2012. Ontogenetic habitat associations of a demersal fish species, \emph{Pagrus auratus}, identified using boosted regression trees. Marine Ecology Progress Series 462:219--230.

\leavevmode\hypertarget{ref-Cramer2009}{}%
Cramer, S. P., and N. K. Ackerman. 2009. Linking stream carrying capacity for salmonids to habitat features. Pages 225--254 \emph{in} American Fisheries Society symposium.

\leavevmode\hypertarget{ref-Dodrill2016}{}%
Dodrill, M. J., and C. B. Yackulic. 2016. Nonlinear relationships can lead to bias in biomass calculations and drift-foraging models when using summaries of invertebrate drift data. Environmental Biology of Fishes 99:659--670.

@@

\leavevmode\hypertarget{ref-Eastwood2003}{}%
Eastwood, P. D., G. J. Meaden, A. Carpentier, and S. I. Rogers. 2003. Estimating limits to the spatial extent and suitability of sole (Solea solea) nursery grounds in the dover strait. Journal of Sea Research 50:151--165.

\leavevmode\hypertarget{ref-Fausch1988}{}%
Fausch, K. D., C. L. Hawkes, and M. G. Parsons. 1988. Models that predict standing crop of stream fish from habitat variables: 1950-85. Notes. General Technical Report, United States Department of Agriculture, Forest Service, Portland, Oregon, USA.

@@

\leavevmode\hypertarget{ref-Hillman1987}{}%
Hillman, T. W., J. Griffith, and W. Platts. 1987. Summer and winter habitat selection by juvenile Chinook salmon in a highly sedimented Idaho stream. Transactions of the American Fisheries Society 116:185--195.

@@

\leavevmode\hypertarget{ref-Holecek2009}{}%
Holecek, D. E., K. J. Cromwell, and B. P. Kennedy. 2009. Juvenile Chinook salmon summer microhabitat availability, use, and selection in a central Idaho wilderness stream. Transactions of the American Fisheries Society 138:633--644.

\leavevmode\hypertarget{ref-Jones2006}{}%
Jones IV, G. P., L. G. Pearlstine, and H. F. Percival. 2006. An assessment of small unmanned aerial vehicles for wildlife research. Wildlife Society Bulletin 34:750--758.
